---
title: "Monocle analysis of influenza-infected cells"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

## Setup
Load or install necessary packages.
If `Monocle` is not already available, install it using `Bioconductor`.
Note that Cole suggested that we might want to install the development version of GitHub; right now this does **not** do this.
```{r, message=FALSE, warning=FALSE}

if (!require("monocle")) {
  source("http://bioconductor.org/biocLite.R")
  biocLite()
  biocLite("monocle")
  }

if (!require("ggExtra")) install.packages("ggExtra")
if (!require("cowplot")) install.packages("cowplot")
```
```{r}
cat("Using Monocle version", as.character(packageVersion("monocle")), "\n")
```

## Create and filter cell-gene dataset
Read in cell-gene data of all samples.
This is the merged cell-gene data set constructed essentially from `cellranger` using the iPython notebook.
The cells are annotated by the number of wildtype and synonymous-barcoded influenza UMIs called for each cell.
```{r}
matrixfile <- "results/cellgenecounts/merged_humanplusflu_matrix.mtx"
genesfile <- "results/cellgenecounts/merged_humanplusflu_genes.tsv"
cellsfile <- "results/cellgenecounts/merged_humanplusflu_cells.tsv"

all_cells <- newCellDataSet(readMM(matrixfile), 
      phenoData=new("AnnotatedDataFrame", data=read.delim(cellsfile)), 
      featureData=new("AnnotatedDataFrame", data=read.delim(genesfile)),
      expressionFamily=negbinomial.size())
```

Filter on just the samples that interest us for this analysis.
```{r}
samples = c("Uninfected", "8hr-2", "8hr-2repeat", "10hr-2")

cellfilter <- row.names(subset(pData(all_cells), Sample %in% samples))
cells <- all_cells[,cellfilter]

cat("Filtered all", nrow(pData(all_cells)), "cells down to the", nrow(pData(cells)), 
      "in samples of interest.")
```

Compute the total number of mRNAs, cellular mRNAs, and influenza mRNAs per cell.
```{r}
pData(cells)$total_mRNAs <- Matrix::colSums(exprs(cells))

flugenes = c("fluPB2", "fluPB1", "fluPA", "fluHA", "fluNP", "fluNA", "fluM", "fluNS")
flu_id = row.names(subset(fData(cells), gene_short_name %in% flugenes))
pData(cells)$flu_mRNAs <- Matrix::colSums(exprs(cells[flu_id,]))

flugenes_id = row.names(subset(fData(cells), !(gene_short_name %in% flugenes)))
pData(cells)$cellular_mRNAs <- Matrix::colSums(exprs(cells[flugenes_id,]))
```

```{r}
scatterplot <- ggplot(pData(cells), aes(cellular_mRNAs, flu_mRNAs)) + geom_point()

# next line uses eval substitute to get value of maxy as here: https://stackoverflow.com/a/10659563
densityplot <- eval(substitute(
    {stat_density(aes(x=cellular_mRNAs, y=maxy*(..scaled..)), geom="line", color='blue')}, 
    list(maxy=layer_scales(scatterplot)$y$range$range[2])))  

scatterplot + densityplot + theme(axis.text.x=element_text(angle=90, vjust=0.5)) + facet_wrap(~Sample)
```
The scatterplots above show the number of influenza mRNAs versus the number of cellular mRNAs for each sample. 
The blue curves show the overall density on the number of cellular mRNAs.
Note that these numbers are not equivalent across samples as the different samples were sequenced to different depths.
The `Monocle` vignette suggests possibly filtering out cells with far more / less mRNAs than most cells.
I have **not** done that here for the reason that this will filter out quite a few of the influenza-infected cells.
This fact is made more clear by this next plot:
```{r}
all_densityplot <- ggplot(pData(cells), aes(total_mRNAs)) + geom_density()
flugt100cells_id <- row.names(subset(pData(cells), flu_mRNAs > 100))
flugt100_densityplot <- ggplot(pData(cells[,flugt100cells_id]), aes(total_mRNAs)) + geom_density()
flugt100_densityplot + facet_wrap(~Sample)
```


```{r}
min_expr <- 5
num_cells <- 5
cells <- detectGenes(cells, min_expr=min_expr)
expressed_genes <- row.names(subset(fData(cells), num_cells_expressed >= num_cells))
cat("Filtering to the", length(expressed_genes), "genes expressed at >=", min_expr,
      "transcripts in at least", num_cells, "cells.\n")
cells <- cells[expressed_genes,]

cat("Plotting number of RNAs per cell for each sample\n")
print(table(pData(cells)$Sample))
pData(cells)$Total_mRNAs <- Matrix::colSums(exprs(cells))
qplot(Total_mRNAs, data=pData(cells), geom="density", color=Sample)

cat("Filtering out 6hr sample since it has a different distribution of RNA per cell")
cells <- cells[, row.names(subset(pData(cells), Sample != "6hr"))]
print(table(pData(cells)$Sample))
```

## Look at distribution of influenza genes.
```{r}
flu_id <- row.names(subset(fData(cells), substr(ShortGeneName, 1, 3) == "flu"))

pData(cells)$Total_flu <- Matrix::colSums(exprs(cells[flu_id,]))
qplot(Total_flu, data=pData(cells), color=Sample, log="x")

cth <- newCellTypeHierarchy()
cth <- addCellType(cth, "noFlu", classify_func=function(x) {colSums(x[flu_id,]) == 0})
cth <- addCellType(cth, "highFlu", classify_func=function(x) {colSums(x[flu_id,]) >= 100}) 
cells <- classifyCells(cells, cth)
print(table(pData(cells)$Sample, pData(cells)$CellType))
```


